Short report: efficacy of vaccination in those eligible for shielding in Scotland

### 2021-08-29 vax2.Rmd
reinfections

freqs.shieldgroups <- as.data.frame.matrix(cbind(
    univariate.tabulate(varnames=c("care.home", "shield.group"), data=cc.vax,
                        drop.reflevel=FALSE),
    univariate.tabulate(varnames=c("care.home", "shield.group"), data=cc.vax[vax14==1],
                        drop.reflevel=FALSE)))[-1, ]
rownames(freqs.shieldgroups) <- table.shieldgroups$effect[9:16]
freqs.shieldgroups$"Crude odds ratio" <- oddsratio.table(freqs.shieldgroups)

freqs.hosp <- as.data.frame.matrix(cbind(
    univariate.tabulate(varnames=c("care.home", "shieldelig.group"), data=cc.hosp.vax,
                        drop.reflevel=FALSE),
    univariate.tabulate(varnames=c("care.home", "shieldelig.group"), data=cc.hosp.vax[vax14==1],
                        drop.reflevel=FALSE)))[-1, ]
rownames(freqs.hosp) <- table.hosp$effect[8:14]
freqs.hosp$"Crude odds ratio" <- oddsratio.table(freqs.hosp)

@@

# Background
In Scotland to date COVD-19 vaccination has been with either the ChAdOx1 nCoV-19 (Astra-Zeneca) or BNT162b2 mRNA (Pfizer-BioNtech) vaccines.  Both have been shown to be efficacious in clinical trials.  Real-world observational analysis in a primary care-based study in Scotland up to 15 Feb 15 2021 found maximal efficacy of 70% for ChAdOx1 nCoV-19 and 85% for BNT162b2 mRNA at 28 days post-vaccination [@vasileiou_effectiveness_2021].  However the efficacy of COVID-19 vaccination among those designated as clinically extremely vulnerable is still uncertain; an official [letter](https://www.gov.scot/publications/coronavirus-covid-19-shielding-letters/) sent on 23 March 2021 to those on this list stated that "we do not yet have evidence of exactly how effective [the vaccine] is for the people on the shielding list."   This is relevant to whether individuals in this category should be advised to remain in shielding as restrictions on the general population are lifted.  This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable and thus eligible for shielding in Scotland.

@@

Vaccination status in cases was coded as having had at least one dose of any vaccine at least 14 days before presentation date. Ths assigns cases (and their matched controls) who were vaccinated less than 14 days prior to the case date to the unexposed category.  The numbers of cases and controls were too few too allow further subdivision of time since vaccination.

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model fitted to the full dataset, specifying effects $\beta_{R2}, \dots, \beta_{RJ}$ for the log rate ratio associated with risk categories 2 to $J$ ($\beta_{R1} = 0$ for the reference category $J=1$), and "nested" effects $\beta_{V1}, \dots, \beta_{VJ}$ for the log rate ratio associated with vaccination in each of the $J$ risk categories.  Because severe COVID-19 is strongly associated with care home residence and care home residents were a priority group for vaccination, the model includes the log rate ratio $\beta_H$ associated with care home residence. For the $i$th individual in a matched set whose care home residence status is $H_i$, is in risk category $j$ and has vaccination status $V_i$, the linear predictor $\Theta_i$ is $\beta_H H_i + \beta_{Rj} + \beta_{Vj} V_{i}$.  The conditional likelihood for a matched set where individual $c$ is the case is $\exp{\left( \Theta_c \right) } / \sum_i{\exp{\left( \Theta_i \right) }}$. The efficacy of vaccination in the $j$th risk category is 1 minus the rate ratio $\exp{\left(\beta_{Vj} \right)}$.

As recent hospital admissions or being in hospital before presentation date might confound the association of vaccination with severe COVID-19, the analysis was repeated with adjustment for any hospital discharge from 2 to 8 weeks before presentation date and for being an inpatient for at least 8 days before presentation date.  Because As another sensitivity analysis and to allow inference on vaccine efficacy in less frequent risk groups, the analysis was repeated with the case definition extended to include all cases that were hospitalized or fatal.  For this analysis the coding of moderate risk conditions was not available, so that those with no risk condition and those with moderate risk conditions are instead grouped together as "ineligible for shielding".

# Results
Figure \ref{fig:plotvax} shows the  proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by presentation date.  This proportion in  controls  will be a close estimate of the  overall  vaccination rates in the total shielding population.  By 9 February 2021 about half the shielded population were vaccinated, rising to over 85% by early March.

@@

Supplementary Table \ref{tab:freqsshieldgroups} shows the distribution of risk groups among  all severe cases and controls and then vaccinated cases and controls. Supplementary Table \ref{tab:freqshosp} shows the same breakdown for hospitalised cases and controls.   These   tables show that the numbers of cases in some shielding subgroups, especially solid organ transplant recipients, are too small for rate ratios to be estimated reliability. Supplementary Table \ref{tab:mhdates} shows the observed and expected numbers of severe cases among vaccinated cases and unvaccinated individuals who eligible for shielding and not resident in care homes, by week of presentation.  This shows that after week 6 of 2021 the number of severe cases was greatere in vaccinated than in unvaccinated individuals.  This is to be expected as by this time more than 70% of those eligible for shielding had been vaccinated; it does not indicate that vaccination was ineffective.

@@

# Conclusions
Efficacy of a single vaccination dose in protecting against severe COVID-19 is as high or higher in those eligible for shielding as in those without risk conditions.  For solid organ transplant recipients -- the group at highest risk among those eligible for shielding -- the number of cases is still too small for the efficacy of vaccination to be estimated.

@@

Since we have previously shown that  the shielding group overall have approximately 6 fold higher risk of severe COVID-19 than those of the same age and sex without risk conditions [@mckeigue_relation_2021-1] they would need a vaccine efficacy of  82% (i.e. a rate ratio of 0.18)  with a single dose to achieve a lowering of risk to the level of unvaccinated persons  without risk conditions.   Thus whilst those in shielding  can be reassured that the vaccine is showing efficacy they should also be advised that they remain at increased risk  after a single dose (about twofold compared to those of the same age and sex without risk conditions given the vaccine-associated rate ratio of about 0.3). Thus they should remain vigilant  until they have received their second dose and background transmission rates are low. Also a  larger dataset than the  Scottish population is needed to estimate efficacy in solid organ transplant recipients where we have not been able to demonstrate efficacy.

The vaccine Green Book recommends that the second dose of all  COVID-19 vaccines should be routinely scheduled between four and 12 weeks after the first dose [@public_health_england_covid-19_2021].  Although earlier rather than later second dosing within that interval might more quickly reduce the residual risk among those in shielding,   this is not necessarily the optimal strategy as the JCVI have noted that "With most vaccines an extended interval between the prime and booster doses leads to a better immune response to the booster dose." [@joint_committee_on_vaccination_and_immunisation_optimising_2021]

@@

\beginsupplement
\onehalfspacing

@@

knitr::kable(freqs.shieldgroups, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqsshieldgroups",
             row.names=TRUE,
             align=c("l", "r", "r", "r", "r"),
             caption="Numbers of controls and cases of severe COVID-19 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "All"=2, "Vaccinated"=2)) %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none") %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="2cm") %>%
    column_spec(4, width="2cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="1cm") %>%
    kable_styling(latex_options=c("hold_position"))

@@

knitr::kable(freqs.hosp, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqshosp",
             row.names=TRUE,
             align=c("l", "r", "r", "r", "r"),
             caption="Numbers of controls and cases of hospitalized or fatal COVID-19 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "All"=2, "Vaccinated"=2)) %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="2cm") %>%
    column_spec(4, width="2cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="1cm") %>%
    kable_styling(latex_options=c("hold_position"))

@@

Vaccine efficacy cannot be estimated from the crude odds ratios in Tables \ref{tab:freqsshieldgroups} and  \ref{tab:freqshosp} because these totals do not take account of the matched case-control design.   The matched design makes it possible to control for age, sex, general practice and crucially for calendar date where the proportion vaccinated and the rates of severe disease were changing day by day during the study period.  The vaccine efficacy is estimated over matched sets, conditional on the marginal totals (cases vs controls, vaccinated versus unvaccinated) in each set. This section is intended to provide a more intuitive explanation of this.  Table \ref{tab:mhdates} shows how information for a statistical test (the Mantel-Haenszel test) of  vaccine efficacy accumulated over matched sets (strata) of cases and controls during the study period.  To keep the working simple, this table is restricted to those not resident in care homes so there is no covariate adjustment.

Each matched set can be represented as a 2 \ensuremath{\times} 2 table, in which the numbers in each cell are traditionally labelled $a, b, c, d$:

 | | Case | Control | All |
 | --- | --- | --- | --- |
 Vaccinated | $a$ | $b$ | $V_1$ |
 Unvaccinated | $c$ | $d$ | $V_0$ |
 All | $C_1$ | $C_0$ | $N$ |

The marginal totals of cases and controls are $C_1, C_0$ and of vaccinated and unvaccinated are $V_1, V_2$.  Conditional on these marginal totals the expected number of vaccinated cases under the null hypothesis of no association is $C_1 V_1 / N$ and the variance of the expected is $C_1 C_0 V_1 V_0 / \left[ N^2 \left(N - 1 \right) \right]$.  The observed numbers, the expected and the variance are summed over matched sets to calculate a test statistic as (observed - expected) divided by the square root of the variance.

@@

>>>>

freqs.shieldgroups <- as.data.frame.matrix(cbind(
   with(cc.vax[CASE==0], paste.rowpercent(table(shield.group, vax14))),
   with(cc.vax[CASE==1], paste.rowpercent(table(shield.group, vax14)))
   ))

freqs.hosp <- as.data.frame.matrix(cbind(
   with(cc.hosp.vax[CASE==0], paste.rowpercent(table(shieldelig.group, vax14))),
   with(cc.hosp.vax[CASE==1], paste.rowpercent(table(shieldelig.group, vax14)))
   ))

rownames(freqs.hosp) <- table.hosp$effect[8:14]

@@

# Key point summary

- This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable and thus eligible for shielding in Scotland.
- This analysis is based on data from 1 December 1st 2020 to 16 March 2021 and therefore covers the initial dose of vaccine with very few second doses in the shielded having been administered as yet  in this period. 
- Efficacy of a single vaccination dose in protecting against severe COVID-19 was found to be as high or higher in those eligible for shielding, considered as a whole,  as in those without risk conditions.   
- Examination of  vaccine efficacy among those eligible for shielding broken down by the reason for shielding showed that for  solid organ transplant recipients the number of cases is still too small for the efficacy of vaccination to be estimated.  For all other groups there was a statistically significant reduction in hospitalisation for COVID-19 associated with  vaccination and that the magnitude of this reduction was similar to that in those who were not eligible for shielding. 
- Further data in larger populations are needed to evaluate vaccine efficacy in solid organ transplant recipients.
- Until such data are available it should not be assumed that vaccination is efficacious in this group.

# Background

In Scotland to date COVD-19 vaccination has been with either the ChAdOx1 nCoV-19 (Astra-Zeneca) or BNT162b2 mRNA (Pfizer-BioNtech) vaccines.  Both have been shown to be efficacious in clinical trials.  Real-world observational analysis in a primary care-based study in Scotland up to 15 Feb 15 2021 found maximal efficacy of 70% for ChAdOx1 nCoV-19 and 85% for BNT162b2 mRNA at 28 days post-vaccination [@vasileiou_effectiveness_2021].  However the efficacy of COVID-19 vaccination among those designated as clinically extremely vulnerable is still uncertain; an official [letter](https://www.gov.scot/publications/coronavirus-covid-19-shielding-letters/) sent on 23 March 2021 to those on this list stated that "we do not yet have evidence of exactly how effective [the vaccine] is for the people on the shielding list."   This is relevant to whether individuals in this category should be advised to remain in shielding as restrictions on the general population are lifted.  This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable and thus eligible for shielding in Scotland.

@@

Vaccination status in cases was coded as having had at least one dose of any vaccine at least 14 days before presentation date. This assigns cases (and their matched controls) who were vaccinated less than 14 days prior to the case date to the unexposed category.  The numbers of cases and controls were too few too allow further subdivision of time since vaccination.

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model fitted to the full dataset, specifying effects $\beta_{R2}, \dots, \beta_{RJ}$ for the log rate ratio associated with risk categories 2 to $J$ ($\beta_{R1} = 0$ for the reference category $J=1$), and "nested" effects $\beta_{V1}, \dots, \beta_{VJ}$ for the log rate ratio associated with vaccination in each of the $J$ risk categories.  Because severe COVID-19 is strongly associated with care home residence and care home residents were a priority group for vaccination, the model includes the log rate ratio $\beta_H$ associated with care home residence. <!---For the $i$th individual in a matched set whose care home residence status is $H_i$, is in risk category $j$ and has vaccination status $V_i$, the linear predictor $\Theta_i$ is $\beta_H H_i + \beta_{Rj} + \beta_{Vj} V_{i}$.  The conditional likelihood for a matched set where individual $c$ is the case is $\exp{\left( \Theta_c \right) } / \sum_i{\exp{\left( \Theta_i \right) }}$. ---> The efficacy of vaccination in the $j$th risk category is 1 minus the rate ratio $\exp{\left(\beta_{Vj} \right)}$.

As recent hospital admissions or being in hospital before presentation date might confound the association of vaccination with severe COVID-19, the analysis was repeated with adjustment for any hospital discharge from 2 to 8 weeks before presentation date and for being an inpatient for at least 8 days before presentation date.   To allow inference on vaccine efficacy in less frequent risk groups, the analysis was also repeated with the case definition extended to include all cases that were hospitalized or fatal.  For this latter analysis the coding of moderate risk conditions was not available, so that those with no risk condition and those with moderate risk conditions are instead grouped together as "ineligible for shielding".

# Results
Figure \ref{fig:plotvax} shows the  proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by presentation date.  This proportion in  controls  will be a close estimate of the  overall  vaccination rates in the total shielding population. From this we can infer that by 9 February 2021 about half the shielded population were vaccinated, rising to over 85% by early March.

@@

Supplementary Table \ref{tab:freqsshieldgroups} shows  the vaccination status of  severe cases and  their controls  by risk stratum.  . Supplementary Table \ref{tab:freqshosp} shows the same breakdown for hospitalised cases and controls.   These   tables show that the numbers of cases in some shielding subgroups, especially solid organ transplant recipients, are too small for rate ratios to be estimated reliability. <!--delete the misleading crude odds ratios from these tables  alter the tables as discussed AND CHANGE THE TITLES ACCORDNGLY-->

@@

For more detail on the computation of the vaccine  efficacy in a matched case control design see  the statistical note in the supplement and the associated  Supplementary Table \ref{tab:mhdates} <!--this comes out as s table  4 but there is no S table 3 please rename--> .

# Conclusions

Efficacy of a single vaccination dose in protecting against severe COVID-19 is as high or higher in those eligible for shielding as in those without risk conditions.  For solid organ transplant recipients -- the group at highest risk among those eligible for shielding -- the number of cases is still too small for the efficacy of vaccination to be estimated.

@@

Since we have previously shown that  the shielding group overall have approximately 6 fold higher risk of severe COVID-19 than those of the same age and sex without risk conditions [@mckeigue_relation_2021-1] they would need a vaccine efficacy of  82% (i.e. a rate ratio of 0.18)  with a single dose to achieve a lowering of risk to the level of unvaccinated persons  without risk conditions.   Thus whilst those in shielding  can be reassured that the vaccine is showing efficacy they should also be advised that they remain at increased risk  after a single dose (about twofold compared to those of the same age and sex without risk conditions given the vaccine-associated rate ratio of about 0.3). Thus they should remain vigilant  until they have received their second dose and background transmission rates are low.

Also a  larger dataset than the  Scottish population is needed to estimate efficacy in solid organ transplant recipients where we have not been able to demonstrate efficacy. It is important that  such an analysis be conducted so that such persons can be advised appropriately as to their ongoing risk level. We note that such data  would also be important to understand the need for development of potential alternatives  to vaccines such as synthetic antibodies that are now being evaluated in clinical trials [@eli_lilly_and_company_phase_2021]

The Vaccine Green Book recommends that the second dose of all  COVID-19 vaccines should be routinely scheduled between four and 12 weeks after the first dose [@public_health_england_covid-19_2021].  Although earlier rather than later second dosing within that interval might more quickly reduce the residual risk among those in shielding,   this is not necessarily the optimal strategy as the JCVI have noted that "With most vaccines an extended interval between the prime and booster doses leads to a better immune response to the booster dose." [@joint_committee_on_vaccination_and_immunisation_optimising_2021]

@@

\beginsupplement
\onehalfspacing
\newpage

@@

knitr::kable(freqs.shieldgroups, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqsshieldgroups",
             row.names=TRUE,
				 col.names=c("Unvaccinated", "Vaccinated", "Unvaccinated", "Vaccinated"),
				 align=c("l", "r", "r", "r", "r"),
             caption="Numbers of controls and cases of severe COVID-19 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "Controls"=2, "Cases"=2)) %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none") %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="2cm") %>%
    column_spec(4, width="2cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="1cm") %>%
    kable_styling(latex_options=c("hold_position"))

@@

knitr::kable(freqs.hosp, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqshosp",
             row.names=TRUE,
  		 col.names=c("Unvaccinated", "Vaccinated", "Unvaccinated", "Vaccinated"),
             align=c("l", "r", "r", "r", "r"),
             caption="Numbers of controls and cases of hospitalized or fatal COVID-19 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "Controls"=2, "Cases"=2)) %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="2cm") %>%
    column_spec(4, width="2cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="1cm") %>%
    kable_styling(latex_options=c("hold_position"))

@@

@@

\newpage

## Statistical note
Tables \ref{tab:freqsshieldgroups} and  \ref{tab:freqshosp}  show the the vaccination status of  distribution of cases and controls by shielding stratum  to illustrate that for solid organ transplant recipients the numbers of cases are too few to estimate vaccine efficacy reliably.   
We have used a matched case-control design, in which each incident case was matched to up to 10 controls of the same age (one-year band), sex and GP practice who were alive on the day of presentation of the case, forming a matched set.  Analyses of these matched sets are thus  inherently stratified by day.  Most matched sets have 1 case and up to 10 controls, but where two cases with the same age, sex and GP practice present on the same day the matched set would have 2 cases and up to 20 controls.

<!---
Each matched set can be represented as a 2 \ensuremath{\times} 2 table, in which the numbers in each cell are traditionally labelled $a, b, c, d$:

 | | Case | Control | All |
 | --- | --- | --- | --- |
 Vaccinated | $a$ | $b$ | $V_1$ |
 Unvaccinated | $c$ | $d$ | $V_0$ |
 All | $C_1$ | $C_0$ | $N$ |
--->
Put simply, the vaccine efficacy is estimated as the most likely value of the odds ratio associated with vaccination given the observed values in each matched set, conditional on the marginal totals (total cases, total controls, total vaccinated, total unvaccinated.  The log-likelihood of the odds ratio given the data is summed across all sets: matched sets in which all or none are vaccinated  do not contribute to the log-likelihood. So for example whilst it is obvious that sets that pertain to days before the vaccination programmes started cannot be informative it may be less obvious that sets pertaining to days when vaccination would be 100% complete are also uninformative. The calendar time period that is most informative is that when around 50% of the population   under study is vaccinated.

When estimating the efficacy within a risk group, only cases and controls in that risk group are retained in the matched sets, and only those only matched sets that contain both unvaccinated and vaccinated individuals contribute to the likelihood.

It follows from the above that crude tables of the vaccination status of cases and controls in each risk group cannot be used to estimate the efficacy.  This is especially true given that vaccination rates are rising sharply over the study period.

Table \ref{tab:mhdates} may give the reader a more intuitive  sense of how the information for computing vaccine efficacy in the shielded evolved over the study period.  It shows the weekly observed split of vaccinated versus unvaccinated cases among those in shielding versus the expected split conditional upon the marginal totals.  To keep the working simple, this table is restricted to those 778 cases eligible for shielding who  were not resident in care homes so there is no covariate adjustment.  The variance of the expected can be interpreted as the information about vaccine efficacy.
